---
title: "Analysis of precipitation trends"
author: "Robert Bagchi"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: pdf_document
number_sections: yes
toc: true
---

```{r packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(tidyselect)
library(reshape2)
library(sf)
library(raster)
library(ggthemes)
library(rnaturalearth)
library(ggspatial)
```

Get precipitation data from Worldclim and coastline from Natural Earth. 

```{r getData}
world <- ne_coastline(scale = "medium", returnclass = "sf")

res <- "2.5"; rcp <- "8.5"
webres <- "2_5"; webrcp <- "85"

bio2000_file <- paste0("../data/wc2.0_", res, "m_bio_1970-2000.zip")
if(!file.exists(bio2000_file))
  download.file(
    paste0(
      "http://biogeo.ucdavis.edu/data/worldclim/v2.0/tif/base/wc2.0_",
      res, "m_bio.zip"), 
    destfile = bio2000_file)
unzip(bio2000_file, files = paste0("wc2.0_bio_", res, "m_12.tif"),  
      overwrite = FALSE, exdir = "../data" )

prec_2000 <- raster(paste0("../data/wc2.0_bio_", res, "m_12.tif"))

gcms <- list("ac", "bc", "cc", "cn", "gf", "gs", "hd", "he", "hg", 
              "in", "ip", "mc","mg", "mi", "mp", "mr", "no")
length(gcms)

sapply(gcms, function(g){
  bio2070_file <- paste0("../data/cmip5/", res, "m/rcp", rcp, "/", 
                         g, rcp, "bi7012.zip")
  if(!file.exists(bio2070_file))
    download.file(
      paste0("http://biogeo.ucdavis.edu/data/climate/cmip5/", webres, "m/",
             g, webrcp,"bi70.zip"), 
      destfile = bio2070_file)
  unzip(bio2070_file, files = paste0(g, webrcp, "bi70", "12.tif"),  
        overwrite = FALSE, 
        exdir = paste0("../data/cmip5/", res, "m/rcp", rcp))
})
lapply(gcms,  function(gcm)
  paste0("../data/cmip5/", res, "m/rcp", rcp, "/", gcm, webrcp,"bi7012.tif"))

```

Organise the data as raster layers.

```{r rasterprep}
prec_2070 <- stack(lapply(gcms,  function(gcm)
  raster(paste0("../data/cmip5/", res, "m/rcp", rcp, "/", 
                gcm, webrcp,"bi7012.tif"))))

prec_2000 <- crop(prec_2000, extent(prec_2070[[1]]))
prec_2070 <-  stack(stack(prec_2000), prec_2070)
names(prec_2070) <- paste0("pr_", c("pres", paste0("fut_", unlist(gcms))))
```

Convert to data frames and then calculate metrics. For the paper, we only present a simple metric from Malhi et al 2008, Science. However, also estimate some other metrics here - the overall patterns are very similar.

```{r calcMetrics}
tropbounds <- list(x=c(-120, 180), y=c(-25, 25))
## get data from with each GCM in a column
tropprec_df <- prec_2070 %>% crop(extent(tropbounds)) %>%
  as.data.frame(xy=TRUE) %>% 
  melt(id.vars=c("x", "y")) %>%  filter(!is.na(value)) %>% 
  spread(key = variable, value = value)

## future columns
futcols <- vars_select(names(tropprec_df), starts_with("pr_fut"))

tropprec_df$pr_fut <- apply(tropprec_df[,futcols], 1, mean) # mean future
tropprec_df$pr_fut_sd <- apply(tropprec_df[,futcols], 1, sd) # sd future

tropprec_df <- tropprec_df %>% mutate(r_pr_fut = pr_fut/pr_pres) ## ratio fut/pres

## get some locations with 0/0, which are no change, so 1
tropprec_df$r_pr_fut[tropprec_df$pr_pres == 0 & tropprec_df$pr_fut==0] <- 1

## z score of change, converted to a probability (experimental).
tropprec_df <- tropprec_df %>% mutate(t_pr_fut = pnorm((pr_fut - pr_pres)/pr_fut_sd))

## set to 0.5 (i.e. no difference) when there is no difference between
## present and future and no variation among future projections.
tropprec_df$t_pr_fut[apply(tropprec_df[, c("pr_pres", futcols)], 
                           1, function(x) var(x)==0)] <- 0.5

## rank of present precip in grid squre compared to projections of 
## future precip in grid squre
## This is the metric used in the paper
tropprec_df$rank_pr_fut <- apply(tropprec_df[,c("pr_pres", futcols)], 1, 
                                 function(x) mean(x[1] < x[-1]))
```

Plotting the projected change in precipitation on a map of the tropics. 

```{r globalmap}
precip_change <-  ggplot() +
  geom_tile(data=filter(tropprec_df, !is.na(t_pr_fut)), 
                          aes(x=x, y=y, fill=t_pr_fut)) +
  geom_sf(data=world, col=NA, fill=NA, size=0.7) +
  coord_sf(xlim = tropbounds$x, ylim = tropbounds$y, expand = FALSE) +
  annotation_scale(location = "bl", width_hint = 0.08) +
  annotation_north_arrow(location = "bl", which_north = "true", 
                         pad_x = unit(0.12, "in"), pad_y = unit(0.25, "in"), 
                         style = north_arrow_nautical) +
  scale_fill_distiller(type="div", direction = 1, limit=c(0, 1)) +
  scale_x_continuous(breaks=seq(-180, 180,20)) +
  scale_y_continuous(breaks=seq(-24, 24, 8)) +
  labs(x="", y="", fill="Pr(Increased Annual \nPrecipitation)") +
  theme_tufte() +  theme(
    panel.grid.major = element_line(color = gray(0.8), linetype = "dashed", 
                                    size = 0.5), 
    panel.background = element_rect(fill = "aliceblue"), 
    plot.margin = unit(c(0.5, 1, 0, 1), "cm"), 
    legend.title=element_text(size=9))

precip_change

ggsave("../figures/world_precip_change_pz2_rcp8.5.png", precip_change,
                 dpi=600, height =4, width=10, units="in")

```
